### 2023-01-22 doc/response.tex

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{xspace}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{chicago}

\newcommand{\eref}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\Rc}{\Rx{\textrm{\tiny{contact}}}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}

\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

\newcommand{\rev}{\subsection*}
\newcommand{\revtext}{\textsf}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0em}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\jsw}[1]{\comment{green}{JSW}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\newcommand{\psymp}{\ensuremath{p}} %% primary symptom time
\newcommand{\ssymp}{\ensuremath{s}} %% secondary symptom time
\newcommand{\pinf}{\ensuremath{\alpha_1}} %% primary infection time
\newcommand{\sinf}{\ensuremath{\alpha_2}} %% secondary infection time

\newcommand{\psize}{{\mathcal P}} %% primary cohort size
\newcommand{\ssize}{{\mathcal S}} %% secondary cohort size

\newcommand{\gtime}{\tau_{\rm g}} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation period distribution

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\begin{document}

\noindent Dear Editor:

Thank you for the chance to revise and resubmit our manuscript.
We have made major revisions to our manuscript to address reviewers' concerns about the methods and assumptions of our work.
We now compare normalized cases divided by the population size---this allows us to directly compare case patterns on campus and those around the surrounding community.
We have reparameterized the community contact rate in our model to account for variation in population sizes across semesters and refitted the model.
We have also added sensitivity analyses and additional figures to show the robustness of our conclusion.
Below please find our detailed responses to reviewers.

\rev{Editor}

\revtext{Associate Editor Comments: }

\revtext{This paper has turned out to be extremely difficult to find sufficient expert reviewers who were willing to review. We have now received two reviews. The first reviewer \#1 is more supportive but the second reviewer \#2 has major concerns that need to be addressed. Please revise the paper to address the concerns and comments of the two reviewers. The revised paper will require further review.}

\revtext{Reviewer \#1 Comments for the Author:}

\revtext{The authors present a retrospective analysis of the COVID-19 epidemic at Princeton University and provide an extensive analysis that generally supports the hypothesis that a superspreading event led to a large uptick in case numbers after the introduction of the Omicron variant. Overall I like the paper. It's well written, and provide a detailed description of the application of a detailed computational model for policy support at an institutional level. It is therefore potentially a valuable contribution and I would support publication after some minor revisions.}

Thank you.

\revtext{Chief among these is the sensitivity of model results (i.e., calibration, and ultimate policy commentary) on the assumptions regarding test sensitivity. The authors claim "Our analysis highlights the power of mass asymptomatic testing" but the assumption is that a PCR test is 95\% sensitive after latency, and that case detection leads to full case isolation. These are both extremely strong assumptions, that are not conservative with respect to the claim. I recommend examining the results of Hellewell et al ("Estimating the effectiveness of routine asymptomatic PCR testing...") for strong evidence that sensitivity is much lower than claimed here. Also see Zachreson et al ("COVID-19 in low-tolerance border quarantine systems...") for a detailed individual-based implementation of dynamic test sensitivity.}

We agree that 95\% sensitivity seem too strong.
Hellewell et al estimated that the probability of detecting an infection from a PCR peaks at 77\% (54–88\%) 4 days after infection, and decreases to 50\% (38–65\%) by 10 days after infection.
But this estimate implicitly accounts for latency, whereas we assume 95\% sensitivity after sensitivity.
Therefore, while our sensitivity peaks at 92\% 5 days after infection, it decreases to 20\% by 10 days after infection (due to recovery from infections), which is considerably lower than what was estimated by Hellewell et al.
Therefore, while the individual-level sensitivity is not dynamic (a step function changing from 0 to 0.95 and back down to 0), the population-level sensitivity is dynamic.
We have added the following paragraph to our main text:

``The 95\% sensitivity assumption may seem too high. 
For example, \cite{hellewell2021estimating} estimated that the probability of detecting an infection from a PCR peaks at 77\% (54–88\%) 4 days after infection, and decreases to 50\% (38–65\%) by 10 days after infection.
These estimates are considerably lower than our assumption because their estimates implicitly account for the latent period.
At the individual level, we assume that an infected individual has no detectable infection (0\% sensitivity) during their latent period and 95\% sensitivity during their various stages of infectious periods.
At the population level, this assumption translates to a peak sensitivity of 92\% by 4 days after infection, decreasing to 20\% sensitivity by 10 days after infection.
Our assumption leads to a much lower PCR sensitivity 10 days after infection because we only model the PCR sensitivity during the infectious period.
In reality, PCR can detect infections even after a person stops shedding infectious virus---we did not include this component in our model because it would not affect the effectiveness of isolation strategy in reducing transmission.''

We further note that the full case isolation was what was happening at Princeton University during the investigation of the study period.
However, we agree that this might be feasible in some institutions. We have added the following phrase:

``while this assumption reflects the isolation policy in Princeton University during the investigation period, it may be inapplicable in studying institutional outbreaks in general.''

\revtext{There are a number of other issues that are less important to my overall recommendation that I list below by section:}

\revtext{Results:}

\revtext{1) Figure 2 (B, E, H) - I would personally prefer to see a representative trajectory (i.e., one that is central in the ensemble based on summary statistics) than a median, since it's being compared to a single real-world trajectory, the 90\% quantiles can stay though. This would give a better illustration of the model dynamics (w.r.t. temporal fluctuations). 
This figure could also benefit from a clearer illustration of which parameters from the grids in (A, D, G) correspond to the ensembles in B, E, H.}

We now show a simulation with the best goodness of fit (least sum of squared errors) alongside a media. 
We agree that showing an actual simulation is important for capturing the variability in the case trajectory.
We also feel that the median is useful for understanding the general trend and decided to keep it in place.
We also added a circle to panels A, D, and G to show the best fitting parameter.

\revtext{2) There is a minor contradiction between Fig 2. A and the statement: "We find that a low level of contacts R contact = 0.5 and a small amount of community transmission $\theta = 0.015$ is most consistent with the observed epidemic dynamics in fall 2020 (Fig. 2A)." in Fig 2A, the minimum sits at Theta = 0.02, and is not definitive over the range > 0.01.}

This was a typo on our side. We have now fixed this.

\revtext{3) it looks like lower values of R could have been explored - what is the reason for the lower bound of 0.5?}

We did this for computationally efficiency. We now explore R as low as 0.25.
We could explore even lower values of R but we feel that lower values may be unrealistic.

\revtext{4) Please explain "Thanksgiving" to non-Americans. This should be indicated visually in Fig. 2B}

We now explain Thanksgiving the first time it is mentioned:

``Thanksgiving---a national holiday in the US during which many students''

\revtext{5) "all returning individuals are assumed to be quarantined for 14 days and tested upon returning." please elaborate (how is quarantine implemented?)}

We have revised this sentence:

``In the beginning of the semester, all returning students were required to quarantine in their rooms for 14 and tested upon returning by the university---in our model, this was implemented by preventing returning students from getting infected or infecting other individuals.''

\revtext{6) Regarding Theta: The model should be refactored so that theta is per-capita on-campus population, the decision not to do so leads to re-scaling of the parameter to campus population fractions in the final section of the results and is confusing. This would probably also help explain why the values of theta are so much higher for the Delta period after vaccination (more people on campus). Finally, the parameter is variously referred to as 'contact rate' and 'transmission rate', this should be clarified in the methods description (they seem to mean the same thing for unvaccinated populations, but the definition gets fuzzier once the transmission event itself is subject to a Bernoulli trial given "infectious contact").}

Thank you for pointing this out. 
We have reparameterized the parameter theta and reran our simulations.
We have also added the following description .

``In particular, we assume that infectious contacts from local or regional community can be made at random to anyone on campus.
These contacts are modeled using a Poisson distribution with a time-varying mean, which is calculated by multiplying the daily number of cases by the community contact rate $\theta$ and the population size on campus $N$.
More precisely, $\theta$ is the probability that an infected individual from the community makes an infectious contact with an individual on campus per capita campus population. 
By further multiplying this probability with the population size $N$, we are essentially assuming a density dependent contact, where a higher population size on campus leads to more infections from the community.''

\revtext{7) The Omicron wave part has a lot of ad-hoc assumptions about policies (i.e., 70 boosters per day), it would be good to know which of these are guessed, and which are supported through consultation etc.}

We have revised this section to make our assumptions clearer:

``Based on known vaccination statuses, we assume that 99\% of students are vaccinated with 60\% of them being boosted as of January 1, 2022.
Since all students were required to receive booster shots before returning to campus, we assumed that 70 booster shots were given on each day---this assumption allows all students to be boosted in 28 days.
To match the high numbers of cases on the week ending January 7, 2022, we assume 14\% of the students present on campus are infected as of January 7, 2022 (roughly 100/700).
To account for students who were infected with the Omicron variant during the fall semester, we assume that 100 students are already immune to Omicron infection at the beginning of the spring semester---this roughly corresponds to the number of PU cases that were reported in December.''

\revtext{8) Figure 3: do the 9 plots on the right have different reproductive ratios? This should be clearly stated (Rcontact is given in the row label, but the column label gives a reproductive ratio multiplier, so it's unclear how assumed transmissibility varies between panels).}

Thank you for pointing this out. The baseline Rcontact does not change within each row; only the reproductive ratio multiplier changes. 
We have added the following sentences to clarify this:

``
For each row, we assume a fixed value of baseline contact reproduction number $\Rc$ ranging from 2 to 6 across rows.
Then, we simulate increase in $\Rc$ at the time of policy change (indicated by column labels).
''

\revtext{9) Please be more clear about the policy change (dashed line in Fig 3). Was this strictly a change to gathering restrictions? Are you implying causation with the super-spreading event? I think the messaging around the relationship between the observed dynamics and the enacted policies needs to be stated more clearly and explicitly.}

There was a change to gathering restriction and testing frequency. This is explained in the Descriptive analysis section:

``Coinciding with the decrease in the campus and local cases numbers, the gathering policy was updated on February 8, 2022 to allow food in events were no longer limited to 20 people;
in addition, the testing frequency was reduced to once a week.''

We have also rephrased the following sentence to make the causality clearer:

``Following the policy change, a large gathering event was held on campus, which resulted in an outbreak with high case numbers persisting until Spring Break (March 5th, 2022).''

\revtext{Discussion:}

\revtext{10) Please explain the phrase "our homogeneous mixing assumption is conservatively pessimistic." One could argue that assuming homogeneity is not 'conservatively pessimistic', rather, it assumes the highest rate of case growth possible for a given natural transmissibility and secondary case dispersion (because transmission is unconstrained by group size beyond the size of the total population), this means the calibrated natural transmission rates (i.e., Rcontact) will be biased down, for the same case incidence. Lower estimates of natural transmission rates (vis. lower viral load of an index case) could lead to liberal conclusions regarding the required stringency of interventions, so it needs to be made clear in what policy decision context the statement 'conservatively pessimistic' applies.}

We meant conservatively pessimistic from a dynamical perspective, allowing for rapid growth.
We agree that this terminology is confusing in a policy decision context. 
We have tried to clarify the consequences our assumption and decided to remove the term ``pessimistic'' to avoid confusion.

``For example, we assume conservatively that the entire university populations mix homogeneously and have identical campus and community contact rates (captured by $\Rc$ and $\theta$, respectively).
This assumption can lead to the fastest epidemic growth rates because transmission is not limited by the size of the contact network---in other words, our estimates of the reproduction will be necessarily low, making the epidemic easier to control.
In reality, increases in cases were often associated with specific transmission clusters, suggesting heterogeneity in transmission patterns.
Contact levels also likely differ between different groups:
for example, faculty and staff members are more likely to interact with community members than undergraduate students and would be at a higher risk for community infections \citep{frazier2022modeling}.''

\revtext{11) The paper claims not to model behaviour. While this is strictly true, because the model does not incorporate behaviour explicitly, it is a bit misleading because the model does implicitly model behaviour by attaching case importation rates to the community case numbers (which are a function of behaviour). A better way of explaining the model would be (in my opinion) that it implicitly accounts for changes in community behaviour but does not explicitly simulate behavioural change in the campus environment.}

Thank you for pointing this out. We have reworded this sentence as the following:

``We also do not account for explicit changes in behavior on campus and assume constant $\Rc$ throughout each semester.
Instead, we implicitly account for behavioral changes in the community by modeling community transmission to campus as a function of community case numbers.
While we cannot rule out the possibility that behavioral changes on campus could have contributed to various epidemics (e.g., the Omicron wave beginning in the fall semester of the 2021--2022 academic year), we were able to capture the majority of epidemic patterns without modeling them---when the majority of transmission is caused by imported cases from the community, we expect behavioral changes on campus to have relatively weaker effects on overall transmission dynamics.''

\revtext{12) "Our analysis highlights the power of mass asymptomatic testing" (see above main comment)}

We have rephrased this sentence:

``First, our analysis highlights the power of mass asymptomatic testing for epidemic measurement and planning---even if PCR testing may have lower sensitivity than what we assumed here \citep{hellewell2021estimating}, mass asymptomatic testing can still help track ongoing epidemic dynamics in real time''

\revtext{13) "preventing large gatherings can help prevent large superspreading events in the midst of a rising epidemic" The model does not simulate gatherings, so this claim should be connected to the results explicitly or it should be removed.}

This was a comment from our Princeton experience rather than simulation. We agree that our model does not simulate gathering and so we have removed this comment.

\revtext{14) "intervention measures placed on campuses must continue to adapt and change" is there a clear statement in this work about the policy changes that led up to the Omicron outbreak? The paper implies such claims, but I do not see them concretely stated (which is frustrating - this is related to point (9)).}

Please see our response to point 9.

\revtext{Methods:}

\revtext{15) Regarding Rcontact, is there a difference between this and the generally-used R0? The definition seems the same... if not, it would be good to know the precise difference.}

Rcontact and R0 are largely similar. But we wanted to use a different terminology to emphasize their subtle differences because Rcontact is a parameter in our individual-based model, which makes a specific set of assumptions that are different from continuous time SIR models using ordinary differential equations. We have tried to clarify the definition of Rcontact.

``We note that the definition of the contact reproduction number $\Rc$ is similar to the standard definition of basic reproduction number $\RR_0$.
The main difference is that the contact reproduction number models the number of total contacts, rather than infections. 
Since infected individuals make their contacts at random with replacement, the same susceptible person could be contacted multiple times by the same or different infected individual during a time step---all these overlapping contacts will result in one infection.
Therefore, the number of actual infections may be smaller than the number of contacts, especially since contacts can also land on non-susceptible individuals.
We also note that the contact reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---therefore, $\Rc$ is similar to the effective reproduction number, which typically accounts for intervention efforts.
However, our contact reproduction number does not account for the effects of asymptomatic testing or vaccination, which are modeled separately.''

\revtext{Overall, I like the paper and appreciate having the opportunity to review it. Understand that my critiques are intended to help improve the quality of the work prior to publication, which I generally support.}

Thank you for your review.

\revtext{Reviewer \#2 Comments for the Author:}

\revtext{In this manuscript, Park et al. retrospectively examine the transmission patterns of COVID-19 at Princeton University and across the wider community in Mercer County, NJ. The authors explore correlations between university and community case counts and also employ a mechanistic modeling framework to see if patterns in transmission can be explained by changes in R-effective and transmission from the broader community. Retrospective analyses like this are important, and valuable evidence generation for decision makers.}

Thank you.

\revtext{I had two main concerns about the conclusions as presented that I hope the authors can address. The first is that while Princeton is an ideal ecosystem because of its asymptomatic screening protocols, the surrounding community is not necessarily one—this raises concerns about how differences in testing practices could be controlled for/accounted for in the correlation analysis.}

\revtext{Most estimates seem to a huge reduction in case ascertainment during the Omicron surge, which makes comparisons between the broader community and Princeton tricky for that time point. At the very least, I would encourage the authors to control for differences in denominator (e.g., converting to per 100K/day) and/or exploring test positivity as an additional metric to see if their observations still hold.}

We have revised our analysis and figures to account for changes in population sizes across the semester. See below for a detailed response.
We also present testing volumes over time and show that they are reasonably stable across each semester.

\revtext{The second is that I wasn’t necessarily convinced that the authors could conclude with certainty that the discrepancies during the Omicron period were driven by superspreading events because the way that this is created in the model is by seeding new infections, not by adjusting contact structure per se. Some more thoughts here: (1) I think the vaccine effectiveness estimates stated for BA.1/BA.2 were overly optimistic, so that would be something to explore further; and (2) the model (in theory) should allow for some superspreading as a consequence of the overdispersion parameter used for the negative binomial distribution of infectious contacts. I’d be curious to see if the model can replicate that change more mechanistically by adjusting the mean and overdispersion of the binomial distribution.}

### 2023-01-22 doc/university-covid.tex
final edits

\begin{flushleft}{
	\Large
	\textbf\newline{
		Relative role of community transmission and campus contagion in driving the spread of SARS-CoV-2: lessons from Princeton University
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1}
Irini Daskalaki\textsuperscript{2}
Robin M. Izzo\textsuperscript{3}
Irina Aranovich\textsuperscript{4}
Aartjan J.W. te Velthuis\textsuperscript{5}
Daniel A. Notterman\textsuperscript{5}
C. Jessica E. Metcalf\textsuperscript{1}
Bryan T.\ Grenfell\textsuperscript{1,6}
\\
\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{2} University Health Services, Princeton University, Princeton, NJ, USA
\\
\textbf{3} Environmental Health and Safety, Princeton University, Princeton, NJ, USA
\\
\textbf{4} Princeton University Clinical Laboratory, Princeton University, Princeton, NJ, USA
\\
\textbf{5} Department of Molecular Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{6} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
\\
\bigskip

@@

Mathematical models have played a crucial role in guiding pandemic responses. 
University campuses present a particularly well-documented case for institutional outbreaks, thereby providing a unique opportunity to understand detailed patterns of pathogen spread.
Here, we present descriptive and modeling analyses of SARS-CoV-2 transmission on the Princeton University campus---this model was used throughout the pandemic to inform policy decisions and operational guidelines for the university campus.
We demonstrate strong spatiotemporal correlations in epidemic patterns between the university campus and surrounding communities.
These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited during much of the wider pandemic until the end of 2021.
Finally, we find that a super-spreading event likely played a major role in driving the recent Omicron variant outbreak on the Princeton University campus during the spring semester of the 2021--2022 academic year.
Despite large numbers of cases on campus in this period, case levels in surrounding communities remained low, suggesting that there was little spillover transmission from campus to the local community.

@@

Mathematical models have played a significant role in guiding these pandemic responses and devising control strategies \citep{cobey2020modeling,holmdahl2020wrong,metcalf2020mathematical,koelle2022changing}.
Models can help monitor key parameters that govern epidemic dynamics \citep{kraemer2021monitoring} and retrospectively estimate the impact of intervention measures in reducing transmission \citep{flaxman2020estimating}.
These estimates can further inform projections of future scenarios and allow us to explore the endemicity of SARS-CoV-2 \citep{kissler2020projecting,saad2020immune,lavine2021immunological,saad2021epidemiological}.
% While models are useful for exploring such counterfactuals, they must be paired with real-life implementations to properly assess the effectiveness of intervention measures.

@@

We begin with a descriptive analysis of the PU outbreak (\fref{princeton}), and present modeling analyses of the individual epidemics during 2020--2022.
PU is located in Mercer County, New Jersey, USA;
the population comprises of 5267 undergraduate students, 2946 graduate students, and around 7000 faculty and staff members.
For simplicity, we divided the epidemic into four time periods representing four semesters across two academic years: Fall 2020--2021 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020--2021 (January 16, 2021--May 14, 2021; \fref{princeton}B), Fall 2021--2022 (August 14, 2021--December 31, 2021; \fref{princeton}C), and Spring 2021--2022 (January 1, 2022--March 18, 2022; \fref{princeton}D).
Throughout the majority of the study period, all students, faculty and staff members who were physically present for more than 8 hours on campus per week were required to participate in asymptomatic testing with varying frequencies.
Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Reverse Transcription Polymerase Chain Reaction (RT-PCR).  
Those who tested positive were required to isolate for at least 10 days after symptom onset or test date (whichever was longer) and were released when they had been at least 48 hours with improving or resolving symptoms as per New Jersey Department of Health guidance.
PCR positives were exempt from asymptomatic testing for 90 days.
Since March 7, 2022, asymptomatic testing frequencies decreased to once a month from once a week for individuals whose vaccine status is up-to-date. This in turn likely reduced the accuracy of surveillance; therefore, we chose to focus on epidemic patterns before this change was implemented.
In Supplementary Figure S1, we show the testing volumes over time.
Throughout the study period, contact tracing was also performed for positive cases to alert their close contacts to either quarantine or test more frequently according to the close contacts' vaccination status, and to gather data that could help uncover clusters of transmission or superspreader events.
Changes in testing frequency and other intervention measures throughout the study period reflected various factors, including the impact of COVID-19 cases on continuity of operations or continuity of teaching; on severity of disease on campus; the capacity of testing and the healthcare system; and hospitalization rates on campus and in the area.
All data used in this analysis are publicly available on the PU COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.

@@

During the \textbf{fall semester of the 2020--2021 academic year}, roughly 1000 grad students and 2000 faculty and staff members were present on campus and participated in asymptomatic testing. 
All classes were held virtually, and so only a few undergraduate students remained on campus ($<300$).  
Both undergraduate and graduate students were required to get tested twice a week, whereas faculty and staff members were required to get tested once a week.
The number of cases remained relatively low throughout the semester with a peak occurring in early December, coinciding with the epidemic trajectory in Mercer County (\fref{princeton}A).  
A sudden decrease in the number of cases around Thanksgiving---a national holiday in the US during which many students---partly reflects the reduced number of tests (3852 and 2972 asymptomatic tests performed on the week ending November 20th and 27th, respectively).
The highest number of cases was reported among faculty and staff members ($=169$), followed by graduate students ($=41$), and undergraduate students ($=4$).
Even when we control for the differing population sizes among these groups, we find that a considerably larger amount of cases were reported among faculty and staff members (84.5 cases per 1000) than from graduate students (41 cases per 1000) (Supplementary Figure S2A)---we exclude undergraduate student population from this comparison due to a very low number of undergraduate students present on campus during this period.

@@

For the \textbf{spring semester of the 2021--2022 academic year}, all eligible students and faculty and staff members were required to be boosted.
By the beginning of the semesters, 65\% of undergraduate students, 71\% of graduate students, and 82\% of faculty and staff members were boosted.
Undergraduate students were still required to be tested twice a week to prevent the additional spread of the Omicron variant.
The number of cases remained high before classes began but decreased over time, following epidemic patterns in Mercer County (\fref{princeton}D).
Coinciding with the decrease in the campus and local cases numbers, the gathering policy was updated on February 8, 2022 to allow food in events were no longer limited to 20 people;
in addition, the testing frequency was reduced to once a week.
Following the policy change, a large gathering event was held on campus, which resulted in an outbreak with high case numbers persisting until Spring Break (March 5th, 2022).
The timing of this outbreak also coincided with a rapid turnover of the Omicron subvariant BA.2---the proportion of BA.2 subvariant reached 93.5\% (372/398) compared to 26.9\% (14/52) from the previous week.
On March 7, 2022, mask mandates were lifted and testing frequency was reduced to once a month.
Cases were largely concentrated among undergraduate students during this semester: 309 cases per 1000 among undergraduate students, 106 cases per 1000 among faculty and staff members, and 65.3 cases per 1000 among graduate students.

@@

Similar picture emerges for Philadelphia except for spring 2020: fall 2020--2021 ($\rho = 0.87$ (95\% CI: 0.68--0.95; $p < 0.001$), Supplementary Figure S6A); spring 2020--2021 ($\rho = 0.27$ (95\% CI: $-0.24$--0.66; $p = 0.30$), Supplementary Figure S6B); fall 2021--2022 ($\rho = 0.89$ (95\% CI: 0.74--0.96; $p < 0.001$), Supplementary Figure S6C); 
and spring 2021--2022 ($\rho=0.46$ (95\% CI: $-0.19$--0.83; $p=0.15$), Supplementary Figure S6D).
Including counties from New York and Pennsylvania states into the spatial correlation analysis yields additional insights (Supplementary Figure S7):
epidemic dynamics were highly synchronized across all counties in fall 2020--2021 and became less synchronized over time. 
These correlations significantly decreased with distance in spring 2020--2021 ($\rho = -0.36$ (95\% CI: $-0.49$--$-0.21$; $p < 0.001$)) and fall 2021--2022 ($\rho = -0.58$ (95\% CI: $-0.68$--$-0.46$; $p < 0.001$)).
These variations likely reflect differences in vaccination levels and the timing of the introduction of the Omicron variant.

>>>>

\begin{flushleft}{
	\Large
	\textbf\newline{
		Relative role of community transmission and campus contagion in driving the spread of SARS-CoV-2: lessons from Princeton University
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1}
Irini Daskalaki\textsuperscript{2}
Robin M. Izzo\textsuperscript{3}
Irina Aranovich\textsuperscript{4}
Aartjan J.W. te Velthuis\textsuperscript{5}
Daniel A. Notterman\textsuperscript{5}
C. Jessica E. Metcalf\textsuperscript{1,6}
Bryan T.\ Grenfell\textsuperscript{1,6}
\\
\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{2} University Health Services, Princeton University, Princeton, NJ, USA
\\
\textbf{3} Environmental Health and Safety, Princeton University, Princeton, NJ, USA
\\
\textbf{4} Princeton University Clinical Laboratory, Princeton University, Princeton, NJ, USA
\\
\textbf{5} Department of Molecular Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{6} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
\\
\bigskip

@@

Mathematical models have played a crucial role in exploring and guiding pandemic responses. 
University campuses present a particularly well-documented case for institutional outbreaks, thereby providing a unique opportunity to understand detailed patterns of pathogen spread.
Here, we present descriptive and modeling analyses of SARS-CoV-2 transmission on the Princeton University campus---this model was used throughout the pandemic to inform policy decisions and operational guidelines for the university campus.
We demonstrate strong spatiotemporal correlations in epidemic patterns between the university campus and surrounding communities.
These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited during much of the wider pandemic until the end of 2021.
Finally, we find that a super-spreading event likely played a major role in driving the Omicron variant outbreak on the Princeton University campus during the spring semester of the 2021--2022 academic year.
Despite large numbers of cases on campus in this period, case levels in surrounding communities remained low, suggesting that there was little spillover transmission from campus to the local community.

@@

Mathematical models have played a significant role in guiding these pandemic responses and exploring control strategies \citep{cobey2020modeling,holmdahl2020wrong,metcalf2020mathematical,koelle2022changing}.
Models can help monitor key parameters that govern epidemic dynamics \citep{kraemer2021monitoring} and retrospectively estimate the impact of intervention measures in reducing transmission \citep{flaxman2020estimating}.
These estimates can further inform projections of future scenarios and allow us to explore the endemicity of SARS-CoV-2 \citep{kissler2020projecting,saad2020immune,lavine2021immunological,saad2021epidemiological}.
% While models are useful for exploring such counterfactuals, they must be paired with real-life implementations to properly assess the effectiveness of intervention measures.

@@

We begin with a descriptive analysis of the PU outbreak (\fref{princeton}), then present modeling analyses of the individual epidemics during 2020--2022.
PU is located in Mercer County, New Jersey, USA;
the population comprises 5267 undergraduate students, 2946 graduate students, and around 7000 faculty and staff members.
For simplicity, we divided the epidemic into four time periods representing four semesters across two academic years: Fall 2020--2021 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020--2021 (January 16, 2021--May 14, 2021; \fref{princeton}B), Fall 2021--2022 (August 14, 2021--December 31, 2021; \fref{princeton}C), and Spring 2021--2022 (January 1, 2022--March 18, 2022; \fref{princeton}D).
Throughout the majority of the study period, all students, faculty and staff members who were physically present for more than 8 hours on campus per week were required to participate in asymptomatic testing with varying frequencies.
Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Reverse Transcription Polymerase Chain Reaction (RT-PCR).  
Those who tested positive were required to isolate for at least 10 days after symptom onset or test date (whichever was longer) and were released when they had been at least 48 hours with improving or resolving symptoms as per New Jersey Department of Health guidance.
PCR positives were exempt from asymptomatic testing for 90 days.
Since March 7, 2022, asymptomatic testing frequencies decreased to once a month from once a week for individuals whose vaccine status is up-to-date. This in turn likely reduced the accuracy of surveillance; therefore, we chose to focus on epidemic patterns before this change was implemented.
In Supplementary Figure S1, we show the testing volumes over time.
Throughout the study period, contact tracing was also performed for positive cases to alert their close contacts to either quarantine or test more frequently according to the close contacts' vaccination status, and to gather data that could help uncover clusters of transmission or superspreader events.
Changes in testing frequency and other intervention measures throughout the study period reflected various factors, including the impact of COVID-19 cases on continuity of operations or continuity of teaching; on severity of disease on campus; the capacity of testing and the healthcare system; and hospitalization rates on campus and in the area.
All data used in this analysis are publicly available on the PU COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.

@@

During the \textbf{fall semester of the 2020--2021 academic year}, roughly 1000 grad students and 2000 faculty and staff members were present on campus and participated in asymptomatic testing. 
All classes were held virtually, and so only a few undergraduate students remained on campus ($<300$).  
Both undergraduate and graduate students were required to get tested twice a week, whereas faculty and staff members were required to get tested once a week.
The number of cases remained relatively low throughout the semester with a peak occurring in early December, coinciding with the epidemic trajectory in Mercer County (\fref{princeton}A).  
A sudden decrease in the number of cases around Thanksgiving---a national holiday in the US during which many students---partly reflects the reduced number of tests (3852 and 2972 asymptomatic tests performed on the week ending November 20th and 27th, respectively).
The highest number of cases was reported among faculty and staff members ($=169$), followed by graduate students ($=41$), and undergraduate students ($=4$).
Even when we control for the differing population sizes among these groups, we find that a considerably larger amount of cases were reported among faculty and staff members (84.5 cases per 1000) than from graduate students (41 cases per 1000) (Supplementary Figure S2A)---we exclude the undergraduate student population from this comparison due to a very low number of undergraduate students present on campus during this period.

@@

For the \textbf{spring semester of the 2021--2022 academic year}, all eligible students and faculty and staff members were required to have obtained booster vaccination.
By the beginning of the semesters, 65\% of undergraduate students, 71\% of graduate students, and 82\% of faculty and staff members were boosted.
Undergraduate students were still required to be tested twice a week to prevent the additional spread of the Omicron variant.
The number of cases remained high before classes began but decreased over time, following epidemic patterns in Mercer County (\fref{princeton}D).
Coinciding with the decrease in the campus and local cases numbers, the gathering policy was updated on February 8, 2022 to allow food in events were no longer limited to 20 people;
in addition, the testing frequency was reduced to once a week.
Following the policy change, a large gathering event was held on campus, which resulted in an outbreak with high case numbers persisting until Spring Break (March 5th, 2022).
The timing of this outbreak also coincided with a rapid increases in the Omicron subvariant BA.2 cases---the proportion of BA.2 subvariant reached 93.5\% (372/398) compared to 26.9\% (14/52) from the previous week.
On March 7, 2022, mask mandates were lifted and testing frequency was reduced to once a month.
Cases were largely concentrated among undergraduate students during this semester: 309 cases per 1000 among undergraduate students, 106 cases per 1000 among faculty and staff members, and 65.3 cases per 1000 among graduate students.

@@

A similar picture emerges for Philadelphia except for spring 2020: fall 2020--2021 ($\rho = 0.87$ (95\% CI: 0.68--0.95; $p < 0.001$), Supplementary Figure S6A); spring 2020--2021 ($\rho = 0.27$ (95\% CI: $-0.24$--0.66; $p = 0.30$), Supplementary Figure S6B); fall 2021--2022 ($\rho = 0.89$ (95\% CI: 0.74--0.96; $p < 0.001$), Supplementary Figure S6C); 
and spring 2021--2022 ($\rho=0.46$ (95\% CI: $-0.19$--0.83; $p=0.15$), Supplementary Figure S6D).
Including counties from New York and Pennsylvania states into the spatial correlation analysis yields additional insights (Supplementary Figure S7):
epidemic dynamics were highly synchronized across all counties in fall 2020--2021 and became less synchronized over time. 
These correlations significantly decreased with distance in spring 2020--2021 ($\rho = -0.36$ (95\% CI: $-0.49$--$-0.21$; $p < 0.001$)) and fall 2021--2022 ($\rho = -0.58$ (95\% CI: $-0.68$--$-0.46$; $p < 0.001$)).
These variations likely reflect differences in vaccination levels and the timing of the introduction of the Omicron variant.
